---
title: "Stat 198 Final Project"
author: 
- "Ursula Brown" 
- "Sara Meltzer" 
- "Anona Thomas"
format: pdf
---

```{r}
#| warning: false
#| echo: false
library(tidyverse)

birth_data <- read.csv("Linked Birth  Infant Death Records, 2017-2023 Expanded.csv") |>
  filter(!is.na(Infant.Birth.Weight.12.Code)) |>
  mutate(
   death_rate = (Deaths / (Deaths + Births)))
```


# Introduction
Infant mortality rate is considered a key indicator of population health. Understanding the factors that contribute to infant death rates is important for improving both maternal and child health outcomes in the United States. In this project, we investigate how month born, infant birth weight, mother's pre-pregnancy BMI, and length of prenatal care influence infant death rate across the US. 

Our primary research question is "To what extent do month of birth, infant's birth weight, mather's pre-pregnancy BMI, and duration of prenatal care influence the likelihood of infant death in the US?" To address this question, we are analyzing a dataset from _______ (someone who knows the website we got it from or the government or whatever put that in)____

To examine the relationship between our selected variables, we are creating a linear regression model. The response variable is infant death rate, and month born, infant birth weight, mother's pre-pregnancy BMI and length of prenatal care are the predictors. In the dataset, infant death rate is a continuous variable. Month born, infant birth weight, mother's pre-pregnancy BMI and length of prenatal care are all categorical variables. Infant birth weight and mother's pre-pregnancy BMI have each been organized into 5 categories, and month prenatal care began has 6 categories. This organization allows for trends to be seen more easily. _____(IS THIS TRUE?)_______(is there a reason its better there's categories?)

This model will give the direction and strength of the association between each predictor and the response variable while controlling for the effects of the other variables in the model. Then, we will be able to answer our research question and determine which predictors influence the likelihood of infant death.

# Methodology

A linear regression model was fit to the available data to determine the influcences that these factors have on the death rate.

```{r}
m1 <- lm(death_rate ~ Month + Infant.Birth.Weight.12 + 
           Mother.s.Pre.pregnancy.BMI + Month.Prenatal.Care.Began, 
         data = birth_data)

summary(m1)
```
Infant birth weight and pre pregnancy BMI are all very significant factors in the death rate!

# Results
```{r}
library(ggplot2)
ggplot(data = birth_data, 
       aes(x = Month, y = death_rate)) +
geom_boxplot()
```
```{r}
ggplot(data = birth_data, 
       aes(x = Infant.Birth.Weight.12, y = death_rate)) +
geom_boxplot()
```
```{r}
ggplot(data = birth_data, 
       aes(x = Mother.s.Pre.pregnancy.BMI, y = death_rate)) +
geom_boxplot()
```
```{r}
ggplot(data = birth_data, 
       aes(x = Month.Prenatal.Care.Began, y = death_rate)) +
geom_boxplot()
```

# Discussion

Overall, we fit a linear regression model to predict the output of infant death rate suing the predictors of month of birth, infant birth weight category, maternal pre-pregnancy BMI, and timing of prenatal care. This model suggests that these variable are highly associated with infant mortality differences with a high predictive value and adjusted R^2 of 0.9842. 

For month of birth, it does not seem to be as significant of a predictor in comparison to the other coefficients in the model. For a a birth occurring in the reference month of April, with the reference birth weight category, reference BMI category (extreme obesity), and prenatal care beginning in the reference month (1st month) is predicted to have an average death rate of 0.130.Compared to the reference month, births in January and November are associated with an average increase in death rate of 0.0117, adjusting for the other variables in the model (p = 0.026).

For maternal pre-pregnancy BMI, most BMI categories (normal, overweight, obese) are associated with slightly lower death rates, with statistically significant effects (p < 0.01 in most cases) when compared to the reference BMI category and adjusting for other variables in the model. Specifically, compared to the reference BMI category, individuals with normal BMI are predicted to have a 0.0217 lower death rate, adjusting for other variables (p < 0.001). 

For pre-natal care, earlier pre-natal care timing is associated with lower infant death rates. Beginning care in the 2nd or 3rd month is associated with a reduction of about 0.055–0.056 in death rate (p < 0.001) when compared to the reference month (the 1st month) and adjusting for all other variables in the model. However, although there is a lower predicted infant mortality with beginning pre-natal care at a month other than the first one, such as the second month, this may be due to confounding variables such as high-risk pregnancies beginning pre-natal care earlier like at month one. 

The strongest association with infant death rate seems to be associated with birth weight. Compared ot the reference category and adjusting for other variables, Extremely low birth weight (<500g) is associated with a much higher death rate (β ≈ +0.375, p < 0.001). Similarly, 500–999g is also strongly associated with higher mortality (β ≈ +0.121, p < 0.001). In contrast, higher birth weight categories (1500g–3999g) are associated with substantially lower death rates compared to the reference group (all p < 0.001). 

Overall, these results suggest that infant birth weight is a very significant predictor of infant mortality in this dataset. Extremely low birth weight infants have dramatically higher death rates, which is consistent with previosu literature (_____CITE_______). Timing of prenatal care also appears important, with earlier care associated with lower mortality, highlighting the benefits of earlier access to emdical care. Maternal BMI shows statistically significant associations, but the effect sizes are small, suggesting it is less influential than birth weight or prenatal care timing. Seasonal variation in birth month appears minimal in this model

This study has several limitations. This analysis can not establish causal relationships, so this model is limited to identifying associations. The data are observational, so there may be confounding variables not included in the model (e.g., socioeconomic status, access to healthcare). Additionally, since this data was collected from (_____ADD WHERE WE GOT FROM___), categories of "unknown" measurements have been included to account for patient privacy and limited data collection, which may detract for the accuracy of this regression model. Additionally, due to the aforementioned reasons, some death rates may be unreliable due to limited reporting. Such limitations indicate that causal relationships should not be drawn.  
